---
title: "Permutation-Based FWER Correction, Part 1: Hypothesis Testing Fundamentals"
author: "Martin Sjogard"
date: 2025-06-20
format: html
categories: [statistics, R, hypothesis testing, neuroscience]
---

## Introduction

In this blog series, we explore permutation-based family-wise error rate (FWER) correction. Before diving into permutation tests and error control, we lay the groundwork by covering core concepts in hypothesis testing: null and alternative hypotheses, test statistics, Type I errors, p-values, and significance testing — all illustrated with hands-on R code.

---

## Null and Alternative Hypotheses

A **statistical hypothesis** is a testable claim about a population or process. We typically test:

- $H_0$: The null hypothesis — no effect or no difference  
- $H_1$: The alternative hypothesis — a meaningful effect or difference

Example:
- $H_0$: A drug does not affect blood pressure
- $H_1$: The drug does affect blood pressure

In neuroscience:
- $H_0$: No difference in brain activity between task and rest
- $H_1$: A significant difference exists

We assume $H_0$ is true and seek evidence to reject it in favor of $H_1$.

---

## Test Statistics and Null Distributions

A **test statistic** is a numeric summary (e.g., a mean difference, $t$-value, or count) used to compare data against $H_0$.

The distribution of the test statistic assuming $H_0$ is true is called the **null distribution**. It provides a reference for assessing how "surprising" the observed statistic is under $H_0$.

---

## Type I Error and Significance Level

There are two main types of errors:

- **Type I error**: Rejecting $H_0$ when it's true (false positive)  
- **Type II error**: Failing to reject $H_0$ when $H_1$ is true (false negative)

The **significance level** $lpha$ (often 0.05) is the acceptable probability of a Type I error.

---

## Interpreting Rejection of $H_0$

- Rejecting $H_0$: Evidence supports $H_1$ (statistically significant)
- Not rejecting $H_0$: Data are consistent with $H_0$; not enough evidence for $H_1$

---

## Understanding p-Values

The **p-value** measures how surprising the observed data are under $H_0$.

- Small p-value → evidence **against** $H_0$
- Large p-value → data are typical under $H_0$

> A p-value is **not** the probability that $H_0$ is true.

---

## R Example: Testing a Coin for Fairness

We test whether a coin is fair using 20 flips.

```{r}
set.seed(42)
n <- 20
flips <- sample(c("H", "T"), size = n, replace = TRUE, prob = c(0.5, 0.5))
num_heads <- sum(flips == "H")
num_heads
test_result <- binom.test(num_heads, n, p = 0.5, alternative = "two.sided")
test_result
```

Interpretation:
- If p > 0.05 → coin is plausibly fair  
- If p ≤ 0.05 → reject $H_0$, suspect bias

---

## Simulated p-Value for Extreme Result

```{r}
# Suppose we observed 17 heads out of 20
p_val <- sum(dbinom(17:20, size = 20, prob = 0.5)) * 2  # two-tailed
p_val
```

---

## Simulation of Null Distribution

```{r}
set.seed(123)
n <- 100
p_null <- 0.5
observed <- 60

# Analytical p-value
p_value_theory <- 1 - pbinom(observed - 1, size = n, prob = p_null)

# Simulated null distribution
reps <- 10000
sim_counts <- rbinom(reps, size = n, prob = p_null)
p_value_sim <- mean(sim_counts >= observed)

# Plot
hist(sim_counts, breaks = 20, col = "skyblue",
     main = "Null Distribution of Heads Count (100 Flips)",
     xlab = "Number of heads")
abline(v = observed, col = "red", lwd = 2)
```

---

## Conclusion

Understanding hypothesis testing, null distributions, and p-values lays the foundation for controlling false positives in multiple comparisons. In the next post, we'll expand to full null distributions, simulation-based inference, and the rationale behind permutation testing.

Stay tuned for **Part 2**.
