---
title: "Permutation-Based FWER Correction, Part 1: Hypothesis Testing Fundamentals"
author: "Martin Sjogard"
date: 2024-06-21
format: html
page: true
categories: [Statistics, Neuroscience, R]
---

I very commonly get questions about nonparametric statistical control from collaborators, and I have been maintaining some personal notes that I tend to distribute to them on request. Here, I have modified them significantly with the intention of making them readable for the younger trainees who also ask me about these things. It should (hopefully) be readable for my collaborators of any background (psychology undergrads, physicians, engineers, neuroscientists).

I will periodically update it with more of my notes. The notes start with establishing some common understanding of hypothesis testing, the concept of a test statistic, distributions of a repeated statistic under the null, and the need for thresholding, then walks (or rambles) through multiple comparisons as a concept, various resampling approaches, some permutation intuition, and finally to the permutation-based family-wise error rate correction and the maximum statistic. The goal is to have the reader think about how distributions are generated under specific constraints and assumptions, and how thresholding these distributions directly relate to the questions we're asking of the data. I find that these permutation-based/non-parametric concepts manage what standard courses on parametric statistics taught in every psychology stats class never does: to instill statistical intuition in people. 

Before diving into permutation tests and error correction, it’s essential to establish a strong foundation in basic statistical inference. In this first post, I will cover the fundamentals of hypothesis testing: what hypotheses are, how we use test statistics, the meaning of Type I errors and significance levels, what it means to “reject $H_0$,” and an intuitive introduction to p-values. I will also walk through a simple example (with R code) to illustrate these concepts in practice.

## Null and Alternative Hypotheses
A **statistical hypothesis** is a claim or conjecture about a population or process that can be tested with data. In hypothesis testing, we usually have two competing hypotheses:

* **Null hypothesis ($H_0$)**: This is the default or status quo hypothesis that there is no effect or no difference. It often represents a baseline assumption or “no change” scenario.

* **Alternative hypothesis ($H_1$ or $H_A$):**: This is the hypothesis that there is an effect, a difference, or a deviation from the null. It typically represents what we are trying to find evidence for.

These two hypotheses are *complementary*: if one is true, the other is false. We design a test to decide whether the data provide sufficient evidence to **reject $H_0$ in favor of $H_1$**. Importantly, we begin by assuming $H_0$ is true (akin to “innocent until proven guilty” in a trial) and then ask whether the data are unlikely enough under that assumption to warrant rejecting it. The alternative hypothesis can be *one-sided* (e.g., an increase in mean, but not a decrease) or *two-sided* (any difference, either an increase or decrease).

**A general example**: Suppose we are testing a new drug.

* $H_0$: The drug has no effect on blood pressure (mean change = 0).

* $H_1$: The drug does have an effect (mean change ≠ 0).

**A neuroimaging example**: In a brain imaging study, we might test if a particular brain region is more active during a task than at rest.

* $H_0$: No difference in brain activity between task and rest in that region.

* $H_1$: A significant difference in activity exists between task and rest.

In both cases, $H_0$ posits “no change/effect,” and $H_1$ posits “some change/effect.” The goal of the test is not to *prove* $H_0$ (we generally assume $H_0$ true until evidence shows otherwise), but rather to determine if we have enough evidence to reject $H_0$ and infer $H_1$.


## Test Statistic
To carry out a hypothesis test, we need a **test statistic** – a numerical summary of the data that we can use to decide between $H_0$ and $H_1$. The test statistic is chosen based on the problem and hypotheses; it should capture the effect we are looking for. Common examples include:

* The difference between two group means (for testing if two groups differ).

* A proportion or count of “successes” (for testing rates or probabilities, like the number of heads in coin flips).

* A correlation coefficient (for testing association between variables).

The key property of a test statistic is that we know (or can determine) its *probability distribution* when $H_0$ is true. This distribution under the null hypothesis is called the **null distribution** of the test statistic. Knowing the null distribution allows us to quantify how extreme our observed statistic is, assuming $H_0$ is correct. For example, if our test statistic is the number of heads in 100 coin flips and $H_0$ states the coin is fair, the null distribution of the statistic is a Binomial(100, 0.5) distribution (the distribution of heads counts for a fair coin).

**Why do we need a test statistic?** It provides a standardized way to evaluate evidence. Raw data can be messy, but a well-chosen statistic (like a mean difference or a $t$-value) condenses the information into a single number that we can compare against a reference distribution. If the observed test statistic falls in a range that would be very unusual under $H_0$ (e.g., far in the tail of the null distribution), it suggests the data are not consistent with $H_0$.

## Type I Error and Significance Level
When making decisions based on data, we must acknowledge the possibility of errors. In hypothesis testing, there are two fundamental error types:

* **Type I error (false positive)**: Rejecting the null hypothesis when it is actually true. In other words, a false alarm – we think we found an effect, but there is none. This is analogous to convicting an innocent person in a trial.

* **Type II error (false negative)**: Failing to reject the null hypothesis when the alternative is true. In other words, a miss – there is a real effect, but our test failed to detect it. Analogously, a guilty person goes free due to lack of evidence.

Because we never have absolute certainty (we rely on samples of data), we cannot eliminate these errors entirely. Instead, we control their probabilities. The **significance level of a test**, denoted $\alpha$, is the probability of making a Type I error that we are willing to tolerate. Common choices are $\alpha = 0.05$ (5%) or $\alpha = 0.01$ (1%). For example, $\alpha = 0.05$ means that if $H_0$ is true, we are willing to accept a 5% chance of mistakenly rejecting it (5% false positive rate). In practical terms, $\alpha=0.05$ means that if $H_0$ is true, the kind of result we deem “significant” would be expected less than 5% of the time by random chance. Note! This is a common way to talk about these values in general but not strictly correct. When I discuss distributions and thresholds more later I will point out the issues with this phrasing.

The significance level defines a **threshold for significance**: if the evidence against $H_0$ is strong enough that the probability of seeing such evidence under $H_0$ is less than $\alpha$, we will reject $H_0$. This threshold can be equivalently thought of in terms of critical values of the test statistic (more on that later) or p-values (coming up next).

It’s important to note that $\alpha$ is set by the researcher *before* seeing the data. This prevents bias in deciding what is “significant.” The smaller $\alpha$ is, the more stringent the test (lower chance of false positive, but higher chance of missing a real effect, i.e., higher Type II error). In the context of multiple comparisons (many tests at once, as often occurs in neuroscience with many brain measurements), controlling the family-wise error rate is essentially about managing the overall Type I error – we’ll get to that in a later post. 

Side note: Continuing the courtroom analogy – if we treat “innocent until proven guilty” as $H_0$, a Type I error is convicting an innocent person, and we set a high bar (stringent $\alpha$) to avoid that. A Type II error is letting a guilty person go free. The significance level is like the “beyond a reasonable doubt” threshold: we decide how much evidence is enough to reject innocence. In hypothesis testing, $\alpha$ is our chosen threshold for doubt. For example, using $\alpha = 0.01$ is like requiring very strong evidence to convict, minimizing false convictions (Type I errors) at the expense of potentially acquitting more truly guilty defendants (some increased Type II errors).


## Rejecting $H_0$: What Does it Mean?
When we say “reject $H_0$”, it means the sample data are sufficiently inconsistent with $H_0$ that we decide $H_0$ is unlikely to be true. In practice, rejecting $H_0$ implies accepting $H_1$ (or at least concluding that there is evidence in favor of $H_1$). A result that leads to rejection is often called “statistically significant” at the chosen $\alpha$ level. This does not mean $H_1$ is proven true in all cases, only that $H_0$ is implausible given the data. We make this decision with a controlled Type I error rate (if $H_0$ is true, we’ll only mistakenly reject it $\alpha$ fraction of the time in repeated experiments).

Conversely, if we “fail to reject $H_0$,” it means the data did not provide strong enough evidence against $H_0$. Important: Failing to reject is not the same as accepting $H_0$ as true – it simply means we do not have sufficient evidence to say it’s false. It’s possible that $H_0$ is false but our sample was not extreme enough (this would be a Type II error). Thus, “not significant” does not prove the null, it only indicates lack of evidence to reject it. 

In summary:

* **Reject $H_0$ (Significant result)**: Data are unlikely under $H_0$ (p-value ≤ α). We infer evidence for $H_1$.

* **Do not reject $H_0$ (Non-significant)**: Data are not sufficiently unusual under $H_0$ (p-value > α). We cannot conclude $H_1$; $H_0$ remains plausible.

This decision process ensures a controlled false positive rate. If $H_0$ is actually true, the chance we incorrectly reject it is α (by design). For example, with α = 0.05, over many repeated experiments with true nulls, about 5% would yield (erroneous) significant results just by chance.

## An Intuitive Introduction to p-Values
Up to now, I’ve talked about “evidence” and outcomes being “unlikely under $H_0$.” The tool that quantifies this is the p-value. We will give a precise definition in the next post, but conceptually: The p-value is a measure of how surprising the observed data would be if the null hypothesis were true. It answers the question: *“If $H_0$ is true, what is the probability of obtaining a result at least as extreme as what we observed?”*

A high p-value (say 0.5) means our result is quite ordinary under $H_0$ (nothing surprising). A very low p-value (say 0.001) means our result would be very rare if $H_0$ were true, indicating that either we saw a fluke or, more plausibly, $H_0$ is false. Some intuitive points about p-values:

* **Small p-value = evidence against $H_0$**: If the data would hardly ever occur under $H_0$, we have reason to doubt $H_0$. For example, getting 95 heads out of 100 coin flips (p-value extremely small under a fair-coin hypothesis) strongly suggests the coin is not fair.

* **Moderate/large p-value = consistency with $H_0$**: If the data are fairly typical under $H_0$, we have no grounds to reject it. E.g., 52 heads out of 100 (p ≈ 0.39 for a fair coin) is quite a common outcome — nothing suspicious.

* **Threshold comparison**: We compare the p-value to α. If p ≤ α, the result is statistically significant (reject $H_0$); if p > α, it’s not significant (fail to reject $H_0$). This is equivalent to the critical-value approach but more directly answers the “how rare is this result under $H_0$?” question.

* **One-tailed vs Two-tailed**: The p-value calculation depends on $H_1$. If $H_1$ is one-sided (e.g., an increase in mean), we calculate the probability of results as extreme as the observed in that direction. If $H_1$ is two-sided (any difference), we consider extremeness in both directions (e.g., both high and low extremes). Two-tailed tests essentially double-count the tail area corresponding to the observed result’s extremeness.

A common misconception is that the p-value is the probability that $H_0$ is true given the data. **This is NOT what the p-value represents**. The p-value is calculated under the assumption that $H_0$ is true; it is *not* the probability of $H_0$ itself. It also isn’t the probability that $H_1$ is true. Instead, it’s about the data: assuming no real effect ($H_0$), how surprising is what we observed? To cement these ideas, let’s walk through a simple example and see how to compute a p-value and make a decision.

## Example: Testing a Coin for Fairness (with R code)

Imagine you have a coin and you suspect it might be biased (not fair). We can frame this as a hypothesis test:

* $H_0$: The coin is fair (probability of heads $p = 0.5$).

* $H_1$: The coin is not fair ($p \neq 0.5$, two-sided alternative).


We decide to flip the coin $n=20$ times and count the number of heads. Our test statistic will be the number of heads in 20 flips. Under $H_0$ (fair coin), the null distribution of this statistic is $\text{Binomial}(n=20, p=0.5)$. Let’s say we perform the experiment. Below is some R code to simulate the coin flips and conduct the test:

```{r}
# Set seed for reproducibility
set.seed(42)

# Define number of flips
n <- 20

# Simulate 20 coin flips (H = "heads", T = "tails")
flips <- sample(c("H", "T"), size = n, replace = TRUE, prob = c(0.5, 0.5))
flips  # show the sequence of flips

# Count the number of heads
num_heads <- sum(flips == "H")
num_heads  # print the number of heads observed

# Perform a two-sided test for fairness:
# H0: p = 0.5, H1: p != 0.5
test_result <- binom.test(num_heads, n, p = 0.5, alternative = "two.sided")
test_result
```

Explanation: I used sample() to simulate 20 flips of a fair coin. I then count heads. The function binom.test() performs an exact binomial test for the null hypothesis $p=0.5$. The output of binom.test will include a p-value. 

If you run this code, you’ll get a certain number of heads (because of randomness, it could be different each run). For instance, one simulation might yield num_heads = 12 out of 20. In that case, the test result will show a p-value (for 12 heads, p ≈ 0.503). This high p-value indicates 12/20 heads is very consistent with a fair coin (no evidence against $H_0$). We would *fail to reject* $H_0$ at α = 0.05. 

Now, to see what a significant result would look like, let’s consider a more extreme outcome. Imagine we had observed num_heads = 17 out of 20. Intuitively, 17 heads in 20 flips is quite unlikely if the coin is fair. We can calculate the p-value for 17 heads:

```{r}
# Compute p-value for observing 17 or more heads out of 20 if p=0.5
p_val <- sum(dbinom(17:20, size = 20, prob = 0.5))
p_val
```

This sums the probabilities of getting 17, 18, 19, or 20 heads under a fair coin (two-tailed test would also include the equally unlikely lower tail: 0–3 heads). The resulting one-tail probability for ≥17 heads is very small (around 0.0013), and doubling for two tails gives ~0.0026. Indeed, binom.test(17, 20, 0.5) would return p ≈ 0.0026. This p-value is *far below* 0.05, so such an outcome would be deemed *highly significant*. We would reject $H_0$ and conclude the coin is likely biased.

To summarize the coin test example:

* If we observe a moderate number of heads (say 8, 10, 12, etc.), the p-value is high (well above 0.05).  We do not have evidence to reject $H_0$; the coin could well be fair.

* If we observe an extreme number of heads (like 17 out of 20), the p-value is very low (≪ 0.05). This is significant evidence against $H_0$; we reject fairness and suspect the coin is biased.

* Our test controlled the Type I error at 5%. If the coin were actually fair, there’s only a 5% chance that we’d see a result as extreme as to falsely conclude bias.


This simple example highlights how hypothesis testing works in practice: define $H_0$ and $H_1$, choose a test statistic and significance level, collect data, compute a p-value (or compare to a critical value), and decide whether or not to reject $H_0$. We used a **p-value approach** here (directly computing the probability of the observed outcome under $H_0$). Alternatively, one could use a **critical value approach**: for instance, for α = 0.05 in the two-sided coin test, the critical region would be ≤3 heads or ≥17 heads (the most extreme 5% of the null distribution). Indeed, 17 heads was the cutoff — that’s why 17 gave p ~0.0026 (half of the 5% two-tailed region). 

## The Null Distribution
Recall that the **null hypothesis ($H_0$)** is the assumption of “no effect” or “no difference.” The **null distribution** is the probability distribution of the test statistic under the assumption that $H_0$ is true. In other words, it tells us how the test statistic would behave just by chance if the null hypothesis holds.

Formally, if $T$ is our test statistic, the null distribution is the distribution of $T$ under $H_0$. For example:

* In the coin flip example, our test statistic was the number of heads in 20 flips. Under $H_0$ (fair coin), $T \sim \text{Binomial}(20, 0.5)$. This binomial distribution is the null distribution of $T$.

* If we test for a difference in means between two large samples under $H_0$ (no true difference), the null distribution of the standardized mean difference might be approximated by a $t$-distribution or normal distribution (depending on assumptions). For instance, in a two-sample *t*-test with equal variances, the null distribution of the *t*-statistic follows a $t$ distribution with appropriate degrees of freedom.

* If testing a correlation’s significance, under $H_0$ (no real correlation), a suitable transform of the correlation coefficient might follow a known distribution (like a t distribution for Pearson correlation with a given sample size).


Why is the null distribution so important? Because it provides the reference frame for significance. We compare our observed test statistic to the null distribution to see *how surprising* the observation is under $H_0$. This comparison is at the heart of calculating p-values and determining critical values. 

Sometimes we can derive the null distribution analytically from probability theory (as with the binomial or t distributions above). However, in many cases the null distribution might be complicated or unknown. In those situations, we often resort to *simulation* or *resampling* techniques (like permutation tests) to empirically estimate the null distribution.

## The Distribution Under $H_0$ vs. Under $H_1$

It’s helpful to contrast the distribution of the test statistic assuming $H_0$ with its distribution assuming a particular alternative $H_1$. Under $H_0$, we know (or estimate) the null distribution. Under $H_1$, the test statistic would typically have a different distribution – often shifted or stretched relative to the null.

For example, consider testing if a coin is biased toward heads. Under $H_0$ (fair coin), the number of heads in 100 flips follows a binomial distribution centered at 50. Under an $H_1$ where the coin is actually biased (say $p=0.6$ chance of heads), the distribution of heads count would center around 60. So under $H_1$, “extreme” high values of heads (say 60 or more) are in fact likely, whereas under $H_0$ they are unlikely. This situation is depicted conceptually as two overlapping distributions: one (null) centered at 50, and another (alternative) centered at 60. A threshold (critical value) set at, say, 58 heads might have a low probability under $H_0$ (thus controlling Type I error) while capturing a lot of the $H_1$ distribution’s mass (thus giving decent power).

* Under $H_0$: the statistic has some known distribution (e.g., centered at 50).

* Under $H_1$: the statistic’s distribution is shifted (e.g., centered at 60 if $p=0.6$).

The *power* of a test is the probability of correctly rejecting $H_0$ when $H_1$ is true – geometrically, it’s the area of the $H_1$ distribution that falls beyond the critical value (in the rejection region). We choose our test/critical value to balance controlling Type I error (false positive rate under $H_0$) and maximizing power (true positive rate under $H_1$). While $H_1$ can encompass many possible values (it’s often a range of scenarios), this framework helps explain why more extreme thresholds reduce false positives but also make it harder to detect real effects.

In practice, we usually focus on the null distribution for calculating p-values and setting up the test, since we assume $H_0$ until evidence indicates otherwise. But it’s good to remember that if there is a real effect ($H_1$ true), the test statistic is expected to deviate from the null distribution systematically.

## Definition and Interpretation of the *p*-value

Now let’s formally define the **p-value**. Given an observed test statistic value (call it $t_{\text{obs}}$), the p-value is the probability of obtaining a test statistic as extreme as (or more extreme than) $t_{\text{obs}}$ *assuming* $H_0$ is true. In formula terms:

* For a right-tailed test (where large values of $T$ favor $H_1$):

$$
\text{p-value} = \Pr_{H_0}\left(T \geq t_{\text{obs}}\right),
$$
the probability that $T$ is greater than or equal to the observed value, under the null distribution.

* For a left-tailed test (where small values favor $H_1$):

$$
\text{p-value} = \Pr_{H_0}\left(T \leq t_{\text{obs}}\right).
$$


* For a two-tailed test (where extreme in either direction counts):

$$
\text{p-value} = \Pr_{H_0}\left(\left|T\right| \geq t_{\text{obs}}\right),
$$
,i.e. the probability of $T$ being as far or farther from the null hypothesis expectation as the observed, in either direction. This can often be computed as twice the one-tailed probability for the tail that $t_{\text{obs}}$ lies in.

In less symbolic language: we locate our observed statistic on the null distribution and see what fraction (or area) of the null distribution lies at least as extreme. The more extreme $t_{\text{obs}}$ is, the smaller this tail area will be, and thus the smaller the p-value. Interpreting the p-value: If the p-value is, say, 0.03, that means “if $H_0$ were true, there is a 3% chance of seeing a result as extreme as this.” A small p-value indicates the observed data are unlikely under $H_0$, which is why we reject $H_0$ when p is below our α threshold. Note that *unlikely under* $H_0$ does not automatically imply *$H_1$ is true* – it simply raises skepticism about $H_0$. (It could be a rare fluke, after all; in the long run, α proportion of true nulls will yield small p-values by chance.) 

To reinforce proper understanding, remember: **The p-value is computed under the assumption that $H_0$ is true**.
. It is not $P(H_0\text{ is true} \mid \text{data})$ and not $P(H_1\text{ is true} \mid \text{data})$. It’s about the data’s compatibility with $H_0$. A low p-value says “this data would be rare if $H_0$ is true,” whereas a high p-value says “this data is quite normal if $H_0$ is true.”

If p < α, we call the result **statistically significant**, and we *reject* $H_0$. If p ≥ α, the result is not significant, and we *do not reject* $H_0$.

For example, at α = 0.05:

* p = 0.001 -> significant (strong evidence against $H_0$),
* p = 0.04 -> significant (evidence against $H_0$),
* p = 0.08 -> not significant (data are not sufficiently unusual under $H_0$).

One more point: When test statistics are continuous, if $H_0$ is true and all assumptions hold, p-values are **uniformly distributed** between 0 and 1. This means that under $H_0$, any p-value is equally likely. As a consequence, if $H_0$ is true, the probability of p ≤ 0.05 is 0.05 (which aligns with our Type I error α). This uniformity is a mathematical fact that ensures our testing procedure is calibrated. If p-values weren’t uniform under $H_0$, a threshold like 0.05 would not correspond exactly to a 5% false positive rate. (This holds for continuous distributions; for discrete cases like the binomial test, p-values have a discrete distribution but still will not be biased toward low values under $H_0$.)


## Critical Values and the Relationship to P-Values
There are two equivalent ways to conduct most hypothesis tests:

* **Critical value approach**: Determine the cutoff value of the test statistic beyond which $H_0$ will be rejected. This cutoff is chosen so that if $H_0$ is true, the probability the test statistic falls beyond the cutoff is α. For instance, in a right-tailed test, the critical value $c$ might be the 95th percentile of the null distribution (so $P_{H_0}(T \ge c) = 0.05$). We reject $H_0$ if the observed $T \ge c$.

* **P-value approach**: Calculate the p-value for the observed test statistic. Reject $H_0$ if p-value ≤ α.

These two approaches lead to the same conclusions. The critical value is essentially the boundary where p-value = α. For example, if the 95th percentile of the null distribution is $c$, then observing $T=c$ yields p-value = 0.05 (for a right-tailed test). Observing $T > c$ yields p < 0.05, and $T < c$ yields p > 0.05.

**Example (critical value vs p-value)**: Suppose $T \sim N(0,1)$ under $H_0$ (a Z-test). For a two-tailed test at α = 0.05, the critical values are approximately ±1.96 (because 2.5% of the distribution is above 1.96 and 2.5% is below -1.96). So:

* Critical region: reject $H_0$ if $T > 1.96$ or $T < -1.96$.

* If our observed $T = 2.5$, this exceeds the critical value, so we reject. The p-value would be $2 \times P(Z > 2.5)$ (two-tailed) which is about 0.0124, indeed ≤ 0.05.

* If observed $T = 1.0$, this does not exceed 1.96, so we fail to reject. The p-value would be $2 \times P(Z > 1.0) ≈ 0.32$, which is > 0.05.


In summary, you can think of the p-value as the *area in the tail beyond the observed statisti*c, and α as the *area in the tail beyond the critical value*. If the observed stat is past the critical value, the tail area beyond it (p-value) will be smaller than α.

## Obtaining Null Distributions: Analytical vs Simulation
To calculate p-values or critical values, we often need the null distribution. There are two main ways to get it:

**1.Analytical derivation (parametric approach)**: Using theoretical probability distributions and assumptions about the data (e.g., normality, independence), we derive the distribution of the test statistic under $H_0$. Classic examples:

* Use the Binomial$(n,p)$ distribution for count of successes under $H_0$.

* Use the $t$ distribution for a t-statistic under $H_0$ (when data are normal or $n$ is large).

* Use the $F$ distribution for an ANOVA F-statistic under $H_0$.

* Use the $\chi^2$ distribution for a chi-square statistic under $H_0$.

These derivations come from probability theory and often involve assumptions (like the data following certain distributions, or large-sample approximations). When conditions are met, the analytical approach is very convenient and fast.

**2. Simulation or resampling (nonparametric approach)**: When the null distribution is complex or unknown, we can simulate it. This could be:

* **Monte Carlo simulation**: Generate many random datasets under $H_0$ (using a known or estimated model) and compute the test statistic for each to see its empirical distribution.

* **Permutation (randomization) tests**: If we have a real dataset, we can create new datasets that satisfy $H_0$ by randomly shuffling or resampling the data. We then calculate the test statistic for each permuted dataset. The distribution of these statistics approximates the null distribution without relying on a formula.


Simulation methods are very useful if, for example, the theoretical null distribution is unknown or too hard to derive, or if we distrust the model assumptions required for the parametric approach. They are computationally heavier but increasingly feasible with modern computing. Permutation tests (a form of simulation) are particularly powerful for preserving the exact data characteristics while enforcing $H_0$ – for instance, shuffling labels in a treatment vs control experiment breaks any real difference (satisfying $H_0$), but keeps the data values themselves intact. I will cover permutation tests in detail later in the series, as they form the basis of permutation-based FWER correction.

For now, I will illustrate obtaining a null distribution by simulation with a simple example in R.

## Example: Generating a Null Distribution via Simulation (and Comparing to Theory)
Consider again the coin-flip scenario, but now with a larger number of flips to get a smoother distribution. Suppose $H_0$: coin is fair ($p=0.5$), and we plan to flip $n=100$ times. Our test statistic is the number of heads in 100 flips. We know analytically that under $H_0$, $T \sim \text{Binomial}(100, 0.5)$.

Let’s obtain the null distribution in two ways: (a) analytical formula, and (b) Monte Carlo simulation of many experiments. I will then use it to compute a p-value for a specific observed outcome (say we observed 60 heads out of 100).

```{r}
set.seed(123)

# Parameters
n <- 100          # number of coin flips
p_null <- 0.5     # probability of heads under H0
observed <- 60    # suppose this is the observed number of heads in the real experiment

# (a) Analytical p-value calculation (one-tailed for illustration: H1 is bias toward heads)
p_value_theory <- 1 - pbinom(observed - 1, size = n, prob = p_null)
p_value_theory  # P(X >= 60) under Binomial(100, 0.5)

# (b) Simulation: generate many experiments under H0
reps <- 10000
sim_counts <- rbinom(reps, size = n, prob = p_null)  # simulate 10,000 experiments
# Empirical p-value from simulation (fraction of experiments with >= 60 heads)
p_value_sim <- mean(sim_counts >= observed)
p_value_sim

# Check that simulation p-value ~ theoretical p-value
abs(p_value_sim - p_value_theory)

# Visualize the null distribution from simulation
hist(sim_counts, breaks = 20, col = "skyblue", main = "Null distribution of heads count (100 flips, p=0.5)",
     xlab = "Number of heads in 100 flips")
abline(v = observed, col = "red", lwd = 2)
```

Let’s break down what I did:

* I set `observed <- 60` as an example outcome (60 heads out of 100 flips).

* The theoretical one-tailed p-value (for the hypothesis of bias towards more heads) is computed as `1 - pbinom(59, 100, 0.5)`. This gives $P(X \ge 60)$ for $X \sim \text{Binomial}(100,0.5)$.

* Finally, I plotted a histogram of the simulated null distribution and drew a red line at the observed value 60.

If you run this, you should see that `p_value_theory` and `p_value_sim` are very close (they’ll differ by a small random error on the order of $10^{-3}$ or so). For instance, you might get something like `p_value_theory ≈ 0.0284` and `p_value_sim ≈ 0.0291` – these are essentially the same, confirming that our simulation is accurately capturing the binomial distribution.

Empirical null distribution of the test statistic (number of heads in 100 flips) under the $H_0$ that the coins are exactly fair: The histogram shows the outcomes of 10,000 simulated fair coin experiments, each with 100 flips. The distribution is roughly bell-shaped, centered around near 50 heads. The red line marks the observed result (60 heads) from a single experiment. The p-value for the observation is the proportion of the null distribution that lies to the right of this red line (the shaded area beyond 60). In this example, the p-value is about 0.03, meaning there is a ~3% chance of seeing 60 or more heads if the coin is truly fair (or, more specifically, if the true expected value of the experiment is 50). 

As expected, the null distribution is centered at 50 (when the coin is fair, on average half the flips are heads). Getting 60 heads is somewhat uncommon under $H_0$ (only ~3% of random fair-coins flips achieved that or more in our simulation). If we were testing $H_0$ at α = 0.05 with the alternative hypothesis “biased toward heads” (one-tailed), we would reject $H_0$ since p ≈ 0.03 < 0.05. If our alternative was two-tailed (detecting any bias), we would consider both tails (≤40 heads or ≥60 heads) which roughly doubles the p-value to ~0.06; in that two-tailed case, 60 heads is actually borderline and would not be significant at the 0.05 level (because while 60 is unusual, so would an equally extreme deficit of heads – our observed outcome is only extreme on the high side).

This example illustrates how we use null distributions to obtain p-values, and how simulation can validate or substitute for analytical calculations. In more complex situations where we can’t write down a formula for the null distribution, we can rely on simulation or permutation to approximate it. The histogram also gives a visual intuition: most outcomes cluster around 50; 60 is out in the tail (hence p ~ 0.03 is the area of that tail).

## The Role of Null Distributions in Multiple Comparisons
*(A brief forward-looking note)*: When we perform **multiple hypothesis tests** (common in fields like neuroimaging, where we often test thousands of voxels for activation differences), the null distribution of the *maximum test statistic* or other aggregate becomes important for controlling the family-wise error rate. Permutation methods often involve simulating the null distribution for the *entire set of tests* by shuffling data labels, and then using the distribution’s quantiles to define corrected critical values that ensure the overall Type I error (FWER) is controlled. In later posts, we will see how the fundamentals from Part 1 and Part 2 come together to address those scenarios. Specifically, we will use permutation-generated null distributions to adjust for multiple comparisons (FWER), ensuring that the chance of any false positive across many tests stays at a desired level (like 5%).

Understanding null distributions and p-values is a critical step in that journey. You should now have a solid grasp of what p-values represent, how to obtain them, and how to interpret them in the context of hypothesis tests. In the next part of this series, I will delve into multiple testing and error rates (including FWER and the false discovery rate) and introduce permutation-based approaches to control these error rates. I’ll show how permutation tests can be applied in a neuroscience context (e.g. analyzing EEG or fMRI data) to make robust inferences while accounting for the multiplicity of comparisons.
