---
title: "Power Analysis for Linear Mixed Models: What You Need to Know"
format: html
author: "Your Name"
date: today
categories: [statistics, R, mixed models, power analysis]
---

## Introduction

Power analysis for linear mixed models (LMMs) is crucial when designing experiments involving repeated measures or hierarchical data structures. In this post, we’ll demonstrate how to use simulation-based methods with the `simr` package in R to estimate power for LMMs.

## Setup

We’ll simulate a dataset mimicking a repeated-measures experiment, fit a mixed model, and then use `simr` for power estimation.

```{r}
# Load required libraries
library(lme4)
library(simr)
library(ggplot2)
```

## Simulate Data

```{r}
set.seed(123)
n_subjects <- 30
n_obs <- 5
subject <- factor(rep(1:n_subjects, each = n_obs))
time <- rep(1:n_obs, n_subjects)
intercept <- 2
slope <- 0.5
subject_effect <- rnorm(n_subjects, 0, 1)
y <- intercept + slope * time + rep(subject_effect, each = n_obs) + rnorm(n_subjects * n_obs, 0, 1)

data <- data.frame(subject, time, y)
```

## Fit the Linear Mixed Model

```{r}
model <- lmer(y ~ time + (1 | subject), data = data)
summary(model)
```

## Power Analysis with `simr`

We extend the model to a larger sample size and compute power for detecting the fixed effect of time.

```{r}
model_extended <- extend(model, along="subject", n=50)
power_result <- powerSim(model_extended, fixed("time", "t"), nsim=100)
power_result
```

## Conclusion

Simulation-based power analysis is an effective approach for complex models like LMMs. The `simr` package makes it straightforward to plan your studies with realistic expectations about effect detectability.

